---
title: "Introduction to RMTnet"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to RMTnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Overview

**RMTnet** builds gene co-expression networks using **Random Matrix Theory (RMT)**
to distinguish real biological signal from finite-sample statistical noise.

## The problem with WGCNA's soft threshold

When you run WGCNA, you must pick a soft-thresholding power β:

```r
# WGCNA — you must eyeball this plot and pick β
powers <- c(1:20)
sft    <- pickSoftThreshold(expr, powerVector = powers)
plot(sft$fitIndices[,1], sft$fitIndices[,2])
# ... pick where R² > 0.8 ... subjective, varies between labs
```

This choice is arbitrary. A different analyst, a different dataset, a different β.
Your network changes substantially based on this one number.

## RMTnet: zero parameters

```{r example, message=FALSE}
library(RMTnet)

# This vignette uses simulated data so it runs anywhere with no external
# dependencies. For results on real data (ALL leukemia), see the README.
mat <- simulate_expression(n_genes = 500, n_samples = 80, n_modules = 3)

# One function call — no parameters to tune
net <- rmt_network(mat)
print(net)
```

The threshold is derived mathematically from your N (genes) and T (samples)
using the **Marchenko-Pastur law** — the same answer every time.

# What is the Marchenko-Pastur law?

Imagine your 500 genes were completely independent — no real biology at all,
just noise. If you still computed their correlation matrix and ran PCA, you
would *still* see non-zero eigenvalues, because finite sample size creates
spurious correlations by chance.

The Marchenko-Pastur law predicts exactly what those spurious eigenvalues
would look like:

$$\rho_{MP}(\lambda) = \frac{Q}{2\pi\sigma^2}
  \frac{\sqrt{(\lambda_+ - \lambda)(\lambda - \lambda_-)}}{\lambda}$$

where $Q = T/N$ and $\lambda_{\pm} = \sigma^2(1 \pm 1/\sqrt{Q})^2$.

Any eigenvalue **above** $\lambda_+$ is statistically significant — more
correlated than random genes could be by chance. Everything below is noise.

```{r spectrum}
# Visualise the eigenvalue spectrum with MP bounds
plot(net, which = "spectrum")
```

# Step-by-step usage

## Step 1: Check your MP threshold

```{r mp}
mp <- mp_threshold(mat)
cat("Genes (N):       ", mp$N, "\n")
cat("Samples (T):     ", mp$T, "\n")
cat("Q = T/N:         ", round(mp$Q, 2), "\n")
cat("MP ceiling λ+:   ", round(mp$lambda_plus, 4), "\n")
cat("Signal PCs:      ", mp$n_signal, "/", mp$N, "\n")
```

## Step 2: Spectral filtering

Noise eigenvalues are replaced by their mean, then the cleaned matrix
is reconstructed. This is analogous to `limma::removeBatchEffect()` but
for statistically-proven noise rather than known confounders.

```{r filter}
filt <- spectral_filter(mp)

# Condition number: lower is better (more stable matrix inversion)
cat("Raw corr condition number:    ",
    round(kappa(filt$raw_corr), 0), "\n")
cat("Cleaned corr condition number:",
    round(kappa(filt$cleaned_corr), 0), "\n")
```

## Step 3: Full network and modules

```{r network}
net <- rmt_network(mat, min_module_size = 20)
net$module_table
```

## Step 4: Visualise

```{r plots, fig.height=8}
plot(net)
```

# Using with real data

RMTnet accepts any normalised expression matrix — the same input you'd give WGCNA:

```r
# From DESeq2
vsd   <- vst(dds)
mat   <- assay(vsd)
net   <- rmt_network(mat)

# From edgeR
logcpm <- cpm(dge, log = TRUE)
net    <- rmt_network(logcpm)

# From a SummarizedExperiment directly
net <- rmt_network(se, assay_name = "logcounts")
```

# Interpreting the output

| Field | What it means |
|---|---|
| `n_signal` | How many PCs are statistically real. Like a rigorous scree plot cutoff. |
| `lambda_plus` | The noise ceiling. Eigenvalues above this are signal. |
| `entropy` | Von Neumann entropy of your expression data. Low = dominated by one factor (batch?). High = diverse. |
| `modules` | Gene module membership, like WGCNA's module colours. |
| `cleaned_corr` | The noise-filtered correlation matrix. Use this for downstream analysis. |

# Session info

```{r sessioninfo}
sessionInfo()
```
